Anomaly detection with kernel density estimation on manifolds

### 2022-04-24 paper/paper.Rmd
Appendix

The rest of the paper is organized as follows.
In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
Conclusions and discussions are presented in \autoref{conclusion}.

@@

## Notations

Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$. The underlying manifold $M$ has the true density $f$. Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$. The Riemannian metric of the embedding is denoted as $g_i, i = 1, 2, \dots, N$. In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$.

@@

Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by $g$ and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold. Denote $A_i, i = 1, 2, \dots, N$, where $A_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with density $f$. For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
\begin{equation}
\label{eq:denriem}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
\end{equation}
where $\theta_p(x)$ is the volume density function on the manifold, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative function, and $h$ is the bandwidth.
@Pelletier2005-vu discussed the volume density function $\theta_p(x)$ defined for $p$ in a neighborhood of $x$ on the manifold, with the geodesic normal coordinates at $x$, $\theta_p(x)$ is equal to the determinant of the Riemannian metric $g$ at $\exp_p^{-1}(x)$. The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].

@@

\clearpage\appendix
\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}
<!-- %\pagenumbering{arabic} -->
<!-- %\def\thepage{A\arabic{page}} -->

<!-- # Notions about Riemannian geometry {#riemgeo} -->

@@

## Canonical metric

@@

For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. 
Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$
<!-- $$ -->
<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
<!-- $$ -->
Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{vol}$ on $T_pM$, and
$\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
$$
d\exp_p^*vol(u) = \theta_p(u)du.
$$

@@

## Laplace-Beltrami operator

>>>>

The rest of the paper is organized as follows.
In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
Conclusions and discussions are presented in \autoref{conclusion}. 
Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.

@@

<!-- ## Notations -->

<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$. The underlying manifold $M$ has the true density $f$. Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$. The Riemannian metric of the embedding is denoted as $g_i, i = 1, 2, \dots, N$. In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$. -->

@@

Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$.
Denote $A_i, i = 1, 2, \dots, N$, where $\pmb{A}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with density $f$. For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
\begin{equation}
\label{eq:denriem}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
\end{equation}
where $h$ is the bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$. The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].

@@

\clearpage\appendix
<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
<!-- %\pagenumbering{arabic} -->
<!-- %\def\thepage{A\arabic{page}} -->

# Appendix: Notions about Riemannian geometry {#riemgeo}

@@

<!-- ## Canonical metric -->

@@

For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. 
Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$
<!-- $$ -->
<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
<!-- $$ -->
Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and
$\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
$$

@@

<!-- ## Laplace-Beltrami operator -->

#### Example

Manifold: S2

$\theta \sim U(0, \pi/2)$

$X = \cos(\theta)$

$$

### 2022-04-24 paper/paper.tex
Appendix

\hypertarget{notations}{%
\subsection{Notations}\label{notations}}

Denote the \(s\)-dimensional input data with observations \(X_i, i = 1, 2, \dots, N\), where \(x_i \in \mathcal{R}^s\). The underlying manifold \(M\) has the true density \(f\). Manifold learning takes the input data and finds its \(d\)-dimensional representation of data that lie on the manifold \(M\), with output data points \(Y_i, i = 1, 2, \dots, N\), where \(Y_i \in \mathcal{R}^d\) and \(d \ll s\). The Riemannian metric of the embedding is denoted as \(g_i, i = 1, 2, \dots, N\). In kernel density estimate, the pointwise variable bandwidth matrix \(H_i\) is used to estimate the density of the embedding, \(\hat{f}\).

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by \(g\) and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold. Denote \(A_i, i = 1, 2, \dots, N\), where \(A_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
\begin{equation}
\label{eq:denriem}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
\end{equation}
where \(\theta_p(x)\) is the volume density function on the manifold, \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative function, and \(h\) is the bandwidth.
\textcite{Pelletier2005-vu} discussed the volume density function \(\theta_p(x)\) defined for \(p\) in a neighborhood of \(x\) on the manifold, with the geodesic normal coordinates at \(x\), \(\theta_p(x)\) is equal to the determinant of the Riemannian metric \(g\) at \(\exp_p^{-1}(x)\). The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.

@@

\clearpage\appendix
\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}

@@

\hypertarget{canonical-metric}{%
\subsection{Canonical metric}\label{canonical-metric}}

@@

\hypertarget{laplace-beltrami-operator}{%
\subsection{Laplace-Beltrami operator}\label{laplace-beltrami-operator}}

>>>>

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\).
Denote \(A_i, i = 1, 2, \dots, N\), where \(\pmb{A}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
\begin{equation}
\label{eq:denriem}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
\end{equation}
where \(h\) is the bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q \(\exp_q^{-1}(p)\). The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.

@@

\clearpage\appendix

\hypertarget{riemgeo}{%
\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}}

@@

@@

\hypertarget{example}{%
\paragraph{Example}\label{example}}

Manifold: S2

\(\theta \sim U(0, \pi/2)\)

\(X = \cos(\theta)\)

\$\$
